# Figure Generation Notebook - SplicingIndex figure

This notebook creates a single figure, SplicingIndexFigure a plot showing the number of AS events per 1000 exons for each chromosome

## Figures (PDF files) created by this notebook
Output PDF Figures are written to the ``pdf/`` directory.

**splicingIndex.pdf**: Plot showing number of tissues-per gene with AS events


```{r}
defaultW <- getOption("warn")  # suppress warnings for this cell
options(warn = -1) 

library(dplyr)
library(ggplot2)
library(multtest)
library(Biobase)
library(tibble)
library(R.utils)
library(rtracklayer)

options(warn = defaultW)
```

### 1. read in gene_as.tsv data containing all significant alternative splicing events

```{r}
filename <- "../data/gene_as.tsv"
gene_as <- read.table(filename, header=TRUE, sep="\t",
                               skipNul=FALSE, stringsAsFactors = FALSE)
head(gene_as,2)
```

### 2 Count most frequent spliced chromosomes
To get an indication of which chromosome has the most frequent slicing event (regardless of type)
We create an index based upon the number of exons per chromosome.

get the annotation file, at this writing, gencode.v30.annotation.gtf
The information as to the number of exons within the chromosome may be found there

```{r}
if (!("gencode.v30.annotation.gtf.gz" %in% list.files("../data/"))) {
    message("downloading gencode v30 annotation\n")
    system("wget -O ../data/gencode.v30.annotation.gtf.gz ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_30/gencode.v30.annotation.gtf.gz")
    message("Done!\n")
    message("Unzipping compressed file gencode.v30.annotation.gtf.gz..")
    system("gunzip ../data/gencode.v30.annotation.gtf.gz", intern = TRUE)
    message("Done! gencode.v30.annotation.gtf can be found in ../data/")
}
gencode <- import("../data/gencode.v30.annotation.gtf")
```

```{r}
exons <- gencode[ gencode$type == "exon", ]
exons <- as.data.frame(exons)

#Obtain chromosomes we have splicing information for (recall we did not use chr Y in our analysis)
all_chr <- as.character(unique(gene_as$chr))
chr_counts <- rep(0, length(all_chr))


for (i in 1:length(all_chr)) {
  chr_counts[i] <- nrow(exons[exons$seqnames == all_chr[i], ])
}

exon_counts <- data.frame(chr = all_chr, counts = chr_counts)

# Count most frequent spliced chromosomes
res <- gene_as %>% group_by(chr) %>% count(chr) %>% arrange(desc(n)) %>% as.data.frame()
res$chr <- factor(res$chr, levels = res$chr)

idx <- match(res$chr, exon_counts$chr)

res$ExonCounts <- exon_counts$counts[idx]

res$Index <- (res$n / res$ExonCounts) * 1000

res_sorted <- res %>% arrange(desc(Index))
res_sorted$chr <- factor(res_sorted$chr, levels = res_sorted$chr)
glimpse(res_sorted)
message("res_sorted has ", length(res_sorted$chr), " chromosomes with data about events, exons, and the splicing index")
```

```{r}
# Colors for uniformity
# Let's use the following colors for uniformity
darkblue <- "#3c5488"
red <- "#e64b35"
nearlyblack <- "#040C04"
purple <- "#790079"
orange = "#ff9900"
```

```{r}
g2 <- ggplot(res_sorted, aes(x = chr, y = Index, size = n)) +
  geom_point(color = darkblue) +
  theme_bw() +
  theme(axis.text.x = element_text(size=16, angle = 270, hjust = 0.0, vjust = 0.5),
        axis.text.y = element_text(size=32),
        axis.title.x = element_blank(),
        axis.title.y = element_text(face="plain", colour="black", 
                                    size=18),
        legend.title=element_text(face="plain", colour="black", 
                                    size=18),
        legend.text = element_text(face="plain", colour="black", 
                                   size=32),
        legend.position = c(0.8, 0.7)) +
  scale_fill_viridis_c(aesthetics = c("colour", "fill"),
                       option = "plasma",
                       limits = c(1, 650)) +
  ylab(paste("Normalized number of sex-biased ASE")) +
  guides(size = guide_legend(title = "Unnormalized\nnumber of\nsex-biased\nASE per\nchromosome"))
ggsave("../pdf/splicingIndex.pdf",g2)
message("Saved plot as ../pdf/splicingIndex.pdf")
g2
```

### Appendix - Metadata

For replicability and reproducibility purposes, we also print the following metadata:

1. Checksums of **'artefacts'**, files generated during the analysis and stored in the folder directory **`data`**
2. List of environment metadata, dependencies, versions of libraries using `utils::sessionInfo()` and [`devtools::session_info()`](https://devtools.r-lib.org/reference/session_info.html)


### Appendix 1. Checksums with the sha256 algorithm

```{r}
notebookid   = "create_SplicingIndex_figure"
notebookid

```

### Appendix 2. Libraries metadata

```{r}
dev_session_info   <- devtools::session_info()
utils_session_info <- utils::sessionInfo()

message("Saving `devtools::session_info()` objects in ../metadata/devtools_session_info.rds  ..")
saveRDS(dev_session_info, file = paste0("../metadata/", notebookid, "_devtools_session_info.rds"))
message("Done!\n")

message("Saving `utils::sessionInfo()` objects in ../metadata/utils_session_info.rds  ..")
saveRDS(utils_session_info, file = paste0("../metadata/", notebookid ,"_utils_info.rds"))
message("Done!\n")

dev_session_info$platform
dev_session_info$packages[dev_session_info$packages$attached==TRUE, ]
```

```{r}


```
