# Figure Generation Notebook - Tissue Piechart figure

This notebook generates a single figure, a tissue piechart figure

## Figures (PDF files) created by this notebook
Output PDF Figures are written to the ``pdf/`` directory.

**tissue-piechart.pdf**: Plot showing number of tissues-per gene with AS events


```{r}
defaultW <- getOption("warn")  # suppress warnings for this cell
options(warn = -1) 

library(dplyr)
library(ggplot2)
library(multtest)
library(Biobase)
library(tibble)
library(R.utils)
library(rtracklayer)

options(warn = defaultW)
```

### 1. Read in all alternatively spliced events

```{r}
filename <- "../data/gene_as.tsv"
gene_as <- read.table(filename, header=TRUE, sep="\t",
                               skipNul=FALSE, stringsAsFactors = FALSE)
head(gene_as,2)
```

### 2. Pie chart
Show recurrent vs unique splicing events with a piechart.


Tissue specific 1 tissue
Tissue group 2-5 tissues
Recurrent > 5 tissues

```{r}
# Colors for uniformity
# Let's use the following colors for uniformity
darkblue <- "#3c5488"
red <- "#e64b35"
nearlyblack <- "#040C04"
purple <- "#790079"
orange = "#ff9900"
```

```{r}
# Count most frequent spliced genes
res <- gene_as %>% group_by(GeneSymbol) %>% count(GeneSymbol) %>% arrange(desc(n)) %>% as.data.frame()
res$GeneSymbol <- factor(res$GeneSymbol, levels = res$GeneSymbol)
length(res$GeneSymbol)

#Add number of tissues
nTissues <- rep(NA, length(res))
for (i in 1:nrow(res)) {
  df_gene <- gene_as %>% filter(GeneSymbol == res$GeneSymbol[i])
  nTissues[i] <- length(unique(df_gene$Tissue))
}
res$Tissues <- nTissues
```

```{r}
head(res)
```

```{r}
#Pie chart - Number of patients with 0, 1, >1 events
counts <- c(res %>% filter(Tissues == 1) %>% count() %>% as.numeric(),
            res %>% filter(Tissues > 1 & Tissues < 6) %>% count() %>% as.numeric(),
            res %>% filter(Tissues > 5) %>% count() %>% as.numeric())
```

```{r}
counts
```

```{r}
piecolors <- c(darkblue,red,orange)
```

Calculate the percentage for each category

```{r}
counts_labels <- round(counts/sum(counts) * 100, 1)
counts_labels <- paste("(",counts_labels,")", "%", sep="")
counts_labels <- paste(counts, counts_labels,sep =" ")
```


```{r}
pdf("../pdf/tissue-piechart.pdf")
pie(counts, main="", col=piecolors, labels=counts_labels, cex=1.5)
# Create a legend at the right   
legend(-.425,0.6, c("1 tissue","2-5 tissues","> 5 tissues"), cex=1.5, 
       fill=piecolors, bg="white")
dev.off()
message("Wrote file to ../pdf/tissue-piechart.pdf")
pie(counts, main="", col=piecolors, labels=counts_labels,
    cex=1.5)
# Create a legend at the right   
legend(-.425,0.6, c("1 tissue","2-5 tissues","> 5 tissues"), cex=1.5, 
       fill=piecolors, bg="white")
```

### Appendix - Metadata

For replicability and reproducibility purposes, we also print the following metadata:

1. Checksums of **'artefacts'**, files generated during the analysis and stored in the folder directory **`data`**
2. List of environment metadata, dependencies, versions of libraries using `utils::sessionInfo()` and [`devtools::session_info()`](https://devtools.r-lib.org/reference/session_info.html)


### Appendix 1. Checksums with the sha256 algorithm

```{r}
notebookid   = "create_tissue_piechart_figure"
notebookid

```

### Appendix 2. Libraries metadata

```{r}
dev_session_info   <- devtools::session_info()
utils_session_info <- utils::sessionInfo()

message("Saving `devtools::session_info()` objects in ../metadata/devtools_session_info.rds  ..")
saveRDS(dev_session_info, file = paste0("../metadata/", notebookid, "_devtools_session_info.rds"))
message("Done!\n")

message("Saving `utils::sessionInfo()` objects in ../metadata/utils_session_info.rds  ..")
saveRDS(utils_session_info, file = paste0("../metadata/", notebookid ,"_utils_info.rds"))
message("Done!\n")

dev_session_info$platform
dev_session_info$packages[dev_session_info$packages$attached==TRUE, ]
```

```{r}

```
