---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.3.3
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# Figure 5  - Two types of exons involved in sex-biased exon skipping. 

- **(a)** log fold change expression vs. log fold change inclusion for all sex-biased events in mammary tissue. A higher fold change of female-to-male expression corresponds to a higher value on the X axis, and a higher fold change of female-to-male skipping-to-inclusion counts on the Y axis. Both fold changes are calculated by limma. Type I events showed a higher frequency of exon skipping in females and type II events showed a higher frequency of inclusion in females. Exon-skipping events are shown in blue instead of red if at least one RBP was significantly associated with the exon-skipping event. The brown lines represent a fit of the points by second-order polynomial regression. 

- **(b)** Comparison of the sum of acceptor and donor
splice scores for type I and type II exons (t-test). 

- **(c)** Comparison of the sum of exon lengths for type I and type II
exons (t-test).


## **NOTE**:

We assume that you have cloned the analysis repository and have `cd` into the parent directory. Before starting with the analysis make sure you have first completed the dependencies set up by following the instructions described in the **`dependencies/README.md`** document. All paths defined in this Notebook are relative to the parent directory (repository). Please close this Notebook and start again by following the above guidelines if you have not completed the aforementioned steps.


# Questions

- [ ] Where can we find this file? `/Users/karleg/Downloads/SplicingCalculatorData/hg38_ensembl_splicing_calculator_output.tsv`
- [ ] Has this file been generated by running a script and if so which one?
- [ ] `meta.data<-read.table('2017December8GTExRNASeqSRARunTable.txt',sep='\t',header=TRUE)` data?
- [ ] `read.table(paste('/Users/karleg/Dimorph/gene_expression/DE_result_',comp,'.txt',sep=''))`
- [ ] `as.tab<-read.table(paste('/Users/karleg/Dimorph/other/',paste(tissue.set,collapse='.'),event.type,'.txt',sep=''))`
- [ ] `event.ensembl.map<-read.table(paste('/Users/karleg/Dimorph/fromGTF.',toupper(event.type),'.txt',sep=''),header=T)`



## Loading dependencies

```{r}
Sys.setenv(TAR = "/bin/tar") # for .gz files

# dataviz dependencies
library(ggplot2)
library(visdat)
library(patchwork)
library(ggsci)
library(grid)
library(report)
library(latex2exp)
```

## 5a. Log-fold change expression vs. log fold change inclusion for all sex-biased events in mammary tissue.

code and `.Rdata` from: [../dimorphAS/figures/figure4c.R](https://github.com/TheJacksonLaboratory/sbas/blob/master/dimorphAS/figures/figure4/Daniels_ScoresHeartLV.R)

```{r}
load("../dimorphAS/figures/figure4/figure4c.RData")
options(repr.plot.width=6.5, repr.plot.height=7)

npgBlue<- rgb(60/256,84/256,136/256,1)
npgRed <- rgb(220/256,0,0,0.5)
npgGreen <- rgb(0,160/256,135/256,1)
npgBrown <- rgb(126/256,97/256,72/256,1)

par(mar=c(5,5,1,1))

plot(x,y,
     main='',
     xlab=TeX("more in males $\\leftarrow$ Differential expression (log FC) $\\rightarrow$ more in females"),
     ylab=TeX("more in males $\\leftarrow$ Exon skipping (log FC) $\\rightarrow$ more in females"),
     cex.lab=1,
     cex.axis=1,
     col=ifelse(as.tab$Row.names[as.tab$Row.names %in% rownames(sig.events)] %in% rbp.regulated$V1,npgBlue,npgRed),
     pch=20)

y.upper<-y[y>0]
x.upper<-x[y>0]
y.lower<-y[y<0]
x.lower<-x[y<0]

fit.upper<-lm(y.upper ~ x.upper + I(x.upper^2))

lines(sort(x.upper), fitted(fit.upper)[order(x.upper)], col=npgBrown, type='l',lwd=1.5) 

fit.lower<-lm(y.lower ~ x.lower + I(x.lower^2))

lines(sort(x.lower), fitted(fit.lower)[order(x.lower)], col=npgBrown, type='l',lwd=1.5) 

abline(v=0,col="black",lty=2,lwd=2)
abline(h=0,col="black",lty=2,lwd=2)

text(1, 3,"Type I" ,cex=1.5)
text(1,-3,"Type II",cex=1.5)
```

## 5b. Comparison of the sum of acceptor and donor splice scores for type I and type II exons (t-test). 

code and `.Rdata` from: [../dimorphAS/figures/figure4/Daniels_ScoresHeartLV.R](https://github.com/TheJacksonLaboratory/sbas/blob/master/dimorphAS/figures/figure4/Daniels_ScoresHeartLV.R)


## Retrieving the required data

```{r}
load('../dimorphAS/figures/figure4/daniels_scoresHeartLV.RData')
options(repr.plot.width=2.5, repr.plot.height=3)

npgBlue<- rgb(60/256,84/256,136/256,1)
npgRed <- rgb(220/256,0,0,0.5)
npgColors <- c(npgBlue,npgRed)

df$type <- as.factor(df$type)  # colors dont work with df$type being numbers

my.pval = t.test(scores.group.1,scores.group.2)$p.value

ggplot(df, aes(x=type, y=score,fill=type, group=type))  +
    
    geom_boxplot(notch=TRUE, 
                 outlier.shape=NA) + 

    theme_minimal()  + 
    scale_y_continuous(limits = quantile(df$score, c(0.1, 0.9))) + 
    guides(fill=FALSE) + theme(axis.text = element_text(size=0, hjust=0.5),
                               axis.ticks.x=element_blank(),
                               axis.title.x=element_blank(), 
                               axis.title.y = element_text(face="bold",  size=12),
                               axis.text.x = element_text(size=12),
                               axis.text.y = element_text(size=12),
                               legend.position="none")   +
    ylab('Donor + Acceptor Score') + 
    annotate("text", 
                    x=1.5, 
                    y=21, 
                    label=  paste0("p==", as.character(formatC(my.pval, 3))),
                    parse=TRUE, 
                    fontface="bold", 
                    size=3) -> pn1
pn1
```

## Get numbers


```{r}
mean(df$score[df$type=="Type I"])
mean(df$score[df$type=="Type II"])
sum(df$type=="Type I")
sum(df$type=="Type II")
```

```{r}
e.lengths.group1<-as.tab1$exonEnd-as.tab1$exonStart_0base
e.lengths.group2<-as.tab2$exonEnd-as.tab2$exonStart_0base

mean(e.lengths.group1)
mean(e.lengths.group2)
```

```{r}
wilcoxon.results <- stats::wilcox.test(e.lengths.group1,
                                       e.lengths.group2,
                                       paired = F)

wilcoxon.results
```

## What does this dataframe do?

```{r}
df<-data.frame(len=c(e.lengths.group1,e.lengths.group2),
               type=c(rep(1,length(e.lengths.group1)),
                      rep(2,length(e.lengths.group2))))

df$type <- ifelse (df$type==1, "Type II", "Type I") 
df$type <- as.factor(df$type) 

head(df)
```

```{r}
my.pval = wilcox.test(e.lengths.group1,e.lengths.group2,paired = F)$p.value
my.pval
```

```{r}
## get numbers
mean(df$score[df$type=="Type I"])
mean(df$score[df$type=="Type II"])
sum(df$type=="Type I")
sum(df$type=="Type II")
e.lengths.group1<-as.tab1$exonEnd-as.tab1$exonStart_0base
e.lengths.group2<-as.tab2$exonEnd-as.tab2$exonStart_0base
mean(e.lengths.group1)
mean(e.lengths.group2)

wilcoxon.results <- wilcox.test(e.lengths.group1,e.lengths.group2,paired = F)


df<-data.frame(len=c(e.lengths.group1,e.lengths.group2),type=c(rep(1,length(e.lengths.group1)),rep(2,length(e.lengths.group2))))
df$type <-  ifelse (df$type==1, "Type II", "Type I") 
df$type <- as.factor(df$type) 

ggplot(df, aes(x=type, 
               y=len,
               fill=type, 
               group=type))  +

            geom_boxplot(notch=TRUE, outlier.shape=NA) + theme_minimal()  +

            scale_y_continuous(limits = quantile(df$len, c(0.1, 0.9))) + guides(fill=FALSE) +

            theme(axis.text = element_text(size=0, hjust=0.5),axis.ticks.x=element_blank(),
                axis.title.x=element_blank(),
                axis.title.y = element_text(face="bold",  size=12),
                axis.text.x = element_text(size=12),
                axis.text.y = element_text(size=12),
                legend.position="none")   + 
            ylab('Exon Length') + 
            annotate("text", 
                     x=1.5, 
                     y=200, 
                     label= as.character(paste0("p = ", formatC(wilcoxon.results$p.value)),
                                         parse=TRUE, 
                                         fontface="bold", 
                                         size=3)) -> pn2
pn2
```

```{r}

options(repr.plot.width=5, repr.plot.height=3)

pn1 + pn2
```

## Metadata

For replicability and reproducibility purposes, we also print the following metadata:

1. Checksums of **'artefacts'**, files generated during the analysis and stored in the folder directory **`data`**
2. List of environment metadata, dependencies, versions of libraries using `utils::sessionInfo()` and [`devtools::session_info()`](https://devtools.r-lib.org/reference/session_info.html)


### 1. Checksums with the sha256 algorithm

```{r}
figure_id   = "figure5"

message("Generating sha256 checksums of the artefacts in the `..data/` directory .. ")
system(paste0("cd ../data/ && sha256sum * > ../metadata/", figure_id, "_sha256sums.txt"), intern = TRUE)
message("Done!\n")

data.table::fread(paste0("../metadata/", figure_id, "_sha256sums.txt"), header = FALSE, col.names = c("sha256sum", "file"))
```

### 2. Libraries metadata

```{r}
figure_id   = "figure5"

dev_session_info   <- devtools::session_info()
utils_session_info <- utils::sessionInfo()

message("Saving `devtools::session_info()` objects in ../metadata/devtools_session_info.rds  ..")
saveRDS(dev_session_info, file = paste0("../metadata/", figure_id, "_devtools_session_info.rds"))
message("Done!\n")

message("Saving `utils::sessionInfo()` objects in ../metadata/utils_session_info.rds  ..")
saveRDS(utils_session_info, file = paste0("../metadata/", figure_id ,"_utils_info.rds"))
message("Done!\n")

dev_session_info$platform
dev_session_info$packages[dev_session_info$packages$attached==TRUE, ]
```
