# Analysis Notebook - summaryAnalysisAnnotations

This notebook looks at specific annotations

```{r}
defaultW <- getOption("warn")  # suppress warnings for this cell
options(warn = -1) 
library(dplyr)
library(R.utils)
library(rtracklayer)

options(warn = defaultW)
```

### 1  TissueReduction Information


```{r}
tissue_reduction_filename <- "../assets/tissues.tsv"
tissue_reduction <- read.table(tissue_reduction_filename, header=TRUE, sep="\t",
                               skipNul=FALSE, stringsAsFactors = FALSE)
colnames(tissue_reduction)  <- c("SMTSD","female","male","include","display_name")
tissue_reduction <- tissue_reduction[tissue_reduction$display_name != "n/a",]
tissue_reduction$display_name <- factor(tissue_reduction$display_name)
levels(tissue_reduction$display_name)
message("We extracted ", length(levels(tissue_reduction$display_name))," different tissues with at least 50 samples in both M & f")
```

```{r}
tissue_reduction <- tissue_reduction[tissue_reduction$include == 1, ]
length(tissue_reduction$display_name)
```

### 2 load the currated srr_pdata 

```{r}
message("Loading srr_pdata\n")
srr_pdata <- readr::read_csv("../data/srr_pdata.csv")
message("done!\n")
```

```{r}
length(unique(srr_pdata$SAMPID))
head(srr_pdata$SAMPID)
length(unique(srr_pdata$SMTSD))
```

### Appendix - Metadata

For replicability and reproducibility purposes, we also print the following metadata:

1. Checksums of **'artefacts'**, files generated during the analysis and stored in the folder directory **`data`**
2. List of environment metadata, dependencies, versions of libraries using `utils::sessionInfo()` and [`devtools::session_info()`](https://devtools.r-lib.org/reference/session_info.html)


### Appendix 1. Checksums with the sha256 algorithm

```{r}
notebookid   = "summaryAnalysisAnnotations"

```

### Appendix 2. Libraries metadata

```{r}
dev_session_info   <- devtools::session_info()
utils_session_info <- utils::sessionInfo()

message("Saving `devtools::session_info()` objects in ../metadata/devtools_session_info.rds  ..")
saveRDS(dev_session_info, file = paste0("../metadata/", notebookid, "_devtools_session_info.rds"))
message("Done!\n")

message("Saving `utils::sessionInfo()` objects in ../metadata/utils_session_info.rds  ..")
saveRDS(utils_session_info, file = paste0("../metadata/", notebookid ,"_utils_info.rds"))
message("Done!\n")

dev_session_info$platform
dev_session_info$packages[dev_session_info$packages$attached==TRUE, ]
```

```{r}

```
