---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.4.1
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# **total Alternative Splicing by Tissue Dot plot**

This notebook generates a single figure, the reverse plot of showing the number of alternative splicing by tissue.

1. Get differential gene expression (DGE) files
2. Use the ``../assets/tissues.tsv`` to limit the tissues to those with at least 50 samples in each sex (``tissues.tsv`` was produced by a Python script)
3. Use the pattern for the differentially expressed genes **"../data/*DGE.csv"** to get all the values for the matrix.


## **Running this notebook**:

See the README for setting up prerequisites for the notebook.


## 1. Setup 

Assumes the `countGenesAndEvents.ipynb` notebook was run -- unpacking the results from the differential Gene Expression Analysis as run in the `differentialGeneExpressionAnalysis.ipynb` notebook.

```{r}
defaultW <- getOption("warn")  # suppress warnings for this cell
options(warn = -1) 

library(stringr)
library(magrittr)
library(dplyr)
library(ggplot2)
library(scales)
library(viridis)

Sys.setenv(TAR = "/bin/tar") # for gzfile

options(warn = defaultW)
```

## 2 read in the total alternatives splicing by tissue file.

```{r}
totals <- read.table("../data/Total_AS_by_tissue.tsv", sep = "\t", header = T)

colnames(totals) <- c("Tissue", "Total")
totals_s <- totals %>% arrange(Total)
totals_s$Tissue <- factor(totals_s$Tissue, levels = totals_s$Tissue)
levels(totals_s$Tissue)
```

## 3 create a function to be used by ggplot to create a reverse log10 scale for the x-axis

```{r}
reverselog_trans <- function(base = exp(1)) {
  trans <- function(x) -log(x, base)
  inv <- function(x) base^(-x)
  trans_new(paste0("reverselog-", format(base)), trans, inv, 
            log_breaks(base = base), 
            domain = c(1e-100, Inf))
}

```

```{r}
g<-ggplot(totals_s, aes(y = Tissue, x = Total, size = Total)) +
  geom_point(color = "red") +
  theme_bw() +
  scale_x_continuous(trans=reverselog_trans(), breaks=c(1,10,100,1000,5000,10000)) +#breaks=c(10000, 5000,1000,100,10,1)) +
  scale_y_discrete(position = "right") +
  theme(axis.text.x = element_text(size=8, angle = 0, hjust = 0.0, vjust = 0.5),
        axis.text.y = element_text(size=8),
        axis.title.x = element_text(face="plain", colour="black", 
                                    size=10),
        axis.title.y = element_blank(),
        legend.title=element_blank(),
        legend.text = element_text(face="plain", colour="black", 
                                   size=8)) +
  xlab(paste("Number of sex-biased splicing events")) +
  ylab("Tissue") + 
  guides(size=FALSE)

g
ggsave("../pdf/total_as_by_tissue.pdf",g, height = 4.5, width = 4)
```

## Appendix - Metadata

For replicability and reproducibility purposes, we also print the following metadata:

### Appendix.1. Checksums with the sha256 algorithm
1. Checksums of **'artefacts'**, files generated during the analysis and stored in the folder directory **`data`**
2. List of environment metadata, dependencies, versions of libraries using `utils::sessionInfo()` and [`devtools::session_info()`](https://devtools.r-lib.org/reference/session_info.html)

```{r}
figure_id   = "totalAlternativeSplicingByTissue"
```

### Appendix.2. Library metadata

```{r}
dev_session_info   <- devtools::session_info()
utils_session_info <- utils::sessionInfo()

message("Saving `devtools::session_info()` objects in ../metadata/devtools_session_info.rds  ..")
saveRDS(dev_session_info, file = paste0("../metadata/", figure_id, "_devtools_session_info.rds"))
message("Done!\n")

message("Saving `utils::sessionInfo()` objects in ../metadata/utils_session_info.rds  ..")
saveRDS(utils_session_info, file = paste0("../metadata/", figure_id ,"_utils_info.rds"))
message("Done!\n")

dev_session_info$platform
dev_session_info$packages[dev_session_info$packages$attached==TRUE, ]
```

```{r}

```
